% !TeX program = tectonic
\documentclass[11pt]{article}
\IfFileExists{alius-guidelines-style.tex}{\input{alius-guidelines-style.tex}}{\input{Guidelines/alius-guidelines-style.tex}}
\renewcommand{\ALIUSFooterTitle}{ALIUS Bulletin Guidelines for Conversations (2022)}

\begin{document}

\guideHeader
  {Conversation}
  {If you want to conduct an open conversation for the ALIUS Bulletin, please read the following guidelines carefully.}
  {Recorded conversation of 30--60 minutes; edited transcript with abstract, keywords, highlights, and bibliography.}
  {Recorded first for flow, then transcribed, edited, approved, and integrated into the annual Bulletin.}
  {\href{mailto:aliusresearchgroup@gmail.com}{aliusresearchgroup@gmail.com}}

\guideSection{Conversation Format}

The conversation will be initially recorded, with the aim of increasing the flow in the exchange of ideas and collective reflection. This will allow interviewers and interviewees to conduct a conversation under naturalistic conditions. The recording does not have to be publicly released in this format. The conversation will then be fully transcribed by an automatic algorithm, and transcripts will finally be edited by both authors and editors. All interviews are formatted and integrated at the end of the year in the annual Bulletin.

\guideSection{Duration of the Conversation}

The recorded conversation is ideally 30--60 minutes long. After transcription and editing, an abstract, keywords, highlights, and bibliography will be added.

\guideSection{Content of the Interview}

You are not expected to carry out the interview in a journalistic style, asking short and non-technical questions. If needed, do not hesitate to ask long and detailed questions. Bear in mind that interviewees might take many things for granted. Therefore, your job, as the interviewer, is to make sure that your questions summarize key concepts and facts in an understandable manner, facilitating the reading for non-specialists.

When relevant, do not hesitate to introduce key concepts and key data to contextualize the questions you are asking. That being said, if you want to ask short questions, feel free to do so: ask long questions only when the intelligibility of what you are saying requires it.

Generally speaking, the interviews are intended for scholars. Therefore, you are not expected to ask general-audience questions as a science journalist would do. However, each interview should be easily understandable to any scholar. For example, an anthropologist reading an interview conducted with a neuroscientist should easily understand most of the discussion, and conversely, a neuroscientist reading an interview conducted with an anthropologist should easily understand most of the discussion.

We invite you to look at the fifth issue of the Bulletin, especially the conversation with Olson and Yaden hosted by George Fejer, to see a concrete example of how questions can be phrased. This model is only an illustration of what can be done. Feel free to differ from it: as long as you broadly follow the above recommendations, it will be perfect.

\guideSection{Modus Operandi and Requirements}

\begin{enumerate}
\item Ask one of the editors of the Bulletin whether the researcher you want to interview fits well with the Bulletin's editorial line.
\item Ask the researcher you want to interview whether they agree in principle to be interviewed. When you do so, let them know about the general concept of the Bulletin, about the other researchers interviewed in previous issues of the Bulletin, and specify that the interview is done in a written way. Describe the modus operandi, the schedule, and so on.
\item You can now write up your questions. Feel free to discuss them with one of the editors if needed, but this is not required.
\item Set up a date to have a recorded conversation with the interviewee, for example via Jitsi or OBS. Explain that the recording will be transcribed into a written format and that nothing will be published without consent to the final edit. Ask the interviewee whether they would like to have the questions in advance.
\item Engage in a conversation with the interviewee. Include an informal opening where you ask the interviewee to say something about their background and/or personality. Ask more specific questions about their work, and include follow-up questions whenever interesting themes arise.
\item Once the conversation is over and recorded, use an automatic algorithm to fully transcribe the interview, or ask the editors for assistance. Upload the recording to the ALIUS Google Drive using the link provided by the editors.
\item Once you have the transcript of the interview, you will next have to:
  \begin{itemize}[label=\textendash,leftmargin=1.5em]
  \item If necessary, compress questions and answers to reduce repetition, overlap, and redundancy.
  \item Edit the transcript to proofread, organize, and shape it into a coherent piece.
  \item Give a title to the interview. It is better to discuss this directly with the interviewee, as you will need their final agreement on the title and the content of the interview.
  \item Write a short abstract, under 200 words, and up to five keywords for referencing the interview. You will need the interviewee's final agreement on these.
  \item If the interviewee discusses books or articles in the interview, insert a bibliography at the end of the interview where all books and articles mentioned by you, in your questions, or by the interviewee, in their responses, are referenced.
  \item Select two to six excerpts, depending on the length of the interview, highlighting key points made by the interviewee and nicely epitomizing their ideas. These passages must be no longer than one or two sentences. Insert the selected excerpts at the very end of the text.
  \end{itemize}
\item When the file is ready, including the interview, title, bibliography, and selected excerpts, send the whole file to the author for approval and make further changes if needed.
\item Once the interviewee and interviewers have agreed on the final version, send the file to the editors. The file must comply with the following requirements:
  \begin{itemize}[label=\textendash,leftmargin=1.5em]
  \item Use Times New Roman, size 12.
  \item Citation style for the bibliography is the 6th edition of APA. If you use Zotero, this citation style can be downloaded from the \href{https://www.zotero.org/styles?q=id%3Aapa}{Zotero APA style page}.
  \item The Bulletin is written in American English rather than British English. If the text that you or the interviewee wrote is in British English, it will have to be turned into American English, for example: programme becomes program, realised becomes realized, and so on. If you or the interviewee uses British lexicon or phrasing, that is fine: you can keep it as it is. Typographic details must still follow American English. Keep in mind that in American English, abbreviations such as ``e.g.'' and ``i.e.'' are followed by a comma, for example: ``i.e., the mind is\ldots'' and not ``i.e. the mind is\ldots''
  \item Use the following quotation marks: `` ''. Do not use guillemets or single quotation marks as the main quotation marks. If there is a quotation within a quotation, use single quotation marks. For example: ``I know that the meaning he gives to `consciousness' differs from\ldots''
  \item Make sure there is no space before punctuation marks such as semicolons, colons, commas, periods, exclamation marks, or question marks.
  \item Dashes used for punctuation must be em dashes and not en dashes, and they must not be separated by spaces. For example: ``the concept of consciousness---thus defined---proves difficult.''
  \item Before sending the file to the editors, take time to carefully go through the interview to make sure there is no typo.
  \end{itemize}
\item Once they receive the text of the interview, the editors will forward it to a native English-speaking proofreader who will check grammar and linguistic correctness. Note that the proofreader's job is not to correct typos. Typos must be corrected before sending the file to the editors.
\item The editors will next send you the feedback provided by the proofreaders. You will be expected to integrate the suggested changes into the interview.
\item Once all changes have been integrated, read the whole interview again very carefully to make sure there is no typo left.
\item When the final version is ready, send it to the interviewee to get their final agreement for publication.
\item Once you have the interviewee's agreement for publication, send the final version of the interview to one of the editors. The interview will then be ready to be published.
\item If the interviewee does not agree to publish the interview as it is and wants changes to be made, make these changes and then send the revised version of the piece to the editors. They will forward it once more to the proofreaders to make sure there are no grammatical or linguistic problems. Once proofreaders have gone through the interview, proceed as described above.
\end{enumerate}

\guideSection{Editorial Team}

For any questions, feel free to ask the editors. ALIUS contact email: \href{mailto:aliusresearchgroup@gmail.com}{aliusresearchgroup@gmail.com}.

Daniel Friedman (\href{mailto:danielarifriedman@gmail.com}{danielarifriedman@gmail.com}); Matthieu Koroma (\href{mailto:matthieu.koroma@gmail.com}{matthieu.koroma@gmail.com}); George Fejer (\href{mailto:gyorgyf@live.com}{gyorgyf@live.com}).

\end{document}
